When environmentally persistent pathogens transform good habitat into ecological traps

### 2016-01-25 Manuscript/manuscript.tex
Formatting citations
Deleting parentheses around citations

\begin{quote}
This draft manuscript is distributed solely for purposes of scientific peer review.  Its content is deliberative and pre decisional, so it must not be disclosed or released by reviewers.  Because the manuscript has not yet been approved for publication by the U.S. Geological Survey (USGS), it does not represent any official USGS finding or policy.
\end{quote}

\newpage

@@

Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Pseudogymnoascus destructans}, \cite{Reynolds2015}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

Since environmental transmission is necessarily a spatial process, its role in disease dynamics may further depend on the spatial structure of the host population.
For many wildlife hosts with environmentally persistent pathogens (e.g. prairie dog colonies affected by plague, \cite{George2013}), this spatial structure is well-described as a metapopulation, where host populations occupy small patches of suitable habitat in a highly fragmented landscape.
Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

We expect that high quality habitat patches, which are more attractive and support greater host density than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants.
The resulting fitness consequences of increased disease risk would then effectively create an ecological trap for migrant hosts preferentially selecting high quality habitat (\cite{Robertson2006}, \cite{Almberg2011}).  
In addition, the greater stability of high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

@@

To address the above questions, we developed a model based on the stochastic patch occupancy model (SPOM) framework.
This framework, based on a well-developed body of theory (\cite{Hanski2003}, \cite{Ovaskainen2004}), models the occupancy state of discrete patches of suitable habitat.
The SPOM approach is flexible and able to capture realistic features of a landscape, including spatial structure and heterogeneity in habitat quality, while remaining relatively tractable (\cite{Ovaskainen2004}).
To incorporate disease processes, we expand on existing SPOM models (\cite{Hanski2003}, \cite{Ovaskainen2004}) and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).  
Transitions between states are governed by host colonization and extinction rates, and a susceptible population can become infected either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).

@@

Continuous time stochastic simulations of the above model were implemented in the R language (\cite{R2014}) using the Gillespie algorithm (\cite{Gillespie1977}).  Code is available at \\
\url{https://github.com/clint-leach/Metapop-Disease}.

@@

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
Moreover, host preference for high quality habitat further exacerbates this increase in disease risk (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
The magnitude of this effect depends on the landscape structure and the scale at which individuals sense and select habitat.
In a fully connected metapopulation, where high quality patches attract migrants from the entire landscape, the effect of preference on the relationship between quality and disease risk is the most pronounced (Supplement Fig 7).
However, the fact that the effect persists in a lattice metapopulation (Supplement Fig 5), where migrants can only select among the four neighboring patches, suggests that host preference for high quality habitat likely serves to further increase the disease risk on high quality patches even in more realistic landscapes.
This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches, and the resulting increased probability of infectious occupancy, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with intermediate spread, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This superspreader role helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists. 
Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability generated by high quality habitat -- the presence of high quality patches serves to substantially increase pathogen spread, even for more virulent pathogens (Figs~\ref{endemic}, \ref{popsizes}).
The consequences of widespread disease-induced mortality can then outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Figs~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction and low colonization rates, are only able to support transient occupancy (Fig~\ref{simvis}, Supplement Figs 4-7), the presence of low quality habitat facilitates more widespread susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. small local population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists. 
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection throughout the metapopulation.
The opportunity for pathogen reservoir decay provided by the ephemeral occupancy of low quality habitat echoes the 'migratory escape' hypothesis (\cite{Loehle1995}), which suggests that annual migration allows hosts to vacate contaminated sites and return once they are clean.
Taken with this literature, our results highlight the importance of transient/temporary habitat in mitigating host exposure to environmentally transmitted pathogens.

@@

These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the potential importance of high-quality habitat for the persistence of wildlife disease.
It is critical to understand these implications in order to avoid falling into the management trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial and should thus be conserved.

The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e. growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e. colonization and extinction).
This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition (\cite{Becker2015}).
Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
More generally, the dynamics and establishment of environmental reservoirs could be influenced by a number of other factors, such as the length of infectious occupancy or the occupancy history of a patch.
The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.
Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g. by making transmission parameters also function of patch quality.

Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of management traps in wildlife populations currently affected by an environmentally persistent pathogen.
For instance, high quality habitat seems unlikely to form management traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of \cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}.
In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a widespread infection, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in moderate disease spread, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the three scenarios we outline here based on disease parameters (infectious survival and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

Given that the presence of this ecological trap depends on the relative magnitude of pathogen spread, we find that pathogens with long-lived environmental reservoirs only generate traps when infectious survival is low enough that the pathogen is not able to spread widely.
Because environmentally long-lived pathogens are able to spread and persist more easily, increasing longevity can overwhelm the ability of low quality habitat to slow spread and provide refuges for susceptible hosts.
These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

>>>>

@@

Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease \cite{Miller2006}, anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Pseudogymnoascus destructans}, \cite{Reynolds2015}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone \cite{Almberg2011, Sharp2011, Breban2009}.

Since environmental transmission is necessarily a spatial process, its role in disease dynamics may further depend on the spatial structure of the host population.
For many wildlife hosts with environmentally persistent pathogens (e.g. prairie dog colonies affected by plague, \cite{George2013}), this spatial structure is well-described as a metapopulation, where host populations occupy small patches of suitable habitat in a highly fragmented landscape.
Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches \cite{Moilanen1998}.  These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

We expect that high quality habitat patches, which are more attractive and support greater host density than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants.
The resulting fitness consequences of increased disease risk would then effectively create an ecological trap for migrant hosts preferentially selecting high quality habitat \cite{Robertson2006, Almberg2011}.  
In addition, the greater stability of high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders \cite{Paull2012}.  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

@@

To address the above questions, we developed a model based on the stochastic patch occupancy model (SPOM) framework.
This framework, based on a well-developed body of theory \cite{Hanski2003, Ovaskainen2004}, models the occupancy state of discrete patches of suitable habitat.
The SPOM approach is flexible and able to capture realistic features of a landscape, including spatial structure and heterogeneity in habitat quality, while remaining relatively tractable \cite{Ovaskainen2004}.
To incorporate disease processes, we expand on existing SPOM models \cite{Hanski2003, Ovaskainen2004} and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).  
Transitions between states are governed by host colonization and extinction rates, and a susceptible population can become infected either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).

@@

Continuous time stochastic simulations of the above model were implemented in the R language \cite{R2014} using the Gillespie algorithm \cite{Gillespie1977}.  Code is available at \\
\url{https://github.com/clint-leach/Metapop-Disease}.

@@

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats \cite{Robertson2006}.
Moreover, host preference for high quality habitat further exacerbates this increase in disease risk (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
The magnitude of this effect depends on the landscape structure and the scale at which individuals sense and select habitat.
In a fully connected metapopulation, where high quality patches attract migrants from the entire landscape, the effect of preference on the relationship between quality and disease risk is the most pronounced (Supplement Fig 7).
However, the fact that the effect persists in a lattice metapopulation (Supplement Fig 5), where migrants can only select among the four neighboring patches, suggests that host preference for high quality habitat likely serves to further increase the disease risk on high quality patches even in more realistic landscapes.
This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected \cite{Christley2005, Keeling2005}.

The relatively low extinction rate on high quality patches, and the resulting increased probability of infectious occupancy, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with intermediate spread, high quality patches are effectively metapopulation scale superspreaders \cite{Lloyd-Smith2005, Paull2012}.  
This superspreader role helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists. 
Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability generated by high quality habitat -- the presence of high quality patches serves to substantially increase pathogen spread, even for more virulent pathogens (Figs~\ref{endemic}, \ref{popsizes}).
The consequences of widespread disease-induced mortality can then outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Figs~\ref{highvlow}, \ref{popsizes}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Figs~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction and low colonization rates, are only able to support transient occupancy (Fig~\ref{simvis}, Supplement Figs 4-7), the presence of low quality habitat facilitates more widespread susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. small local population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists. 
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection throughout the metapopulation.
The opportunity for pathogen reservoir decay provided by the ephemeral occupancy of low quality habitat echoes the 'migratory escape' hypothesis \cite{Loehle1995}, which suggests that annual migration allows hosts to vacate contaminated sites and return once they are clean.
Taken with this literature, our results highlight the importance of transient/temporary habitat in mitigating host exposure to environmentally transmitted pathogens.

@@

These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the potential importance of high-quality habitat for the persistence of wildlife disease.
It is critical to understand these implications in order to avoid falling into the management trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e. high quality habitat), but Strasser et al. \cite{Strasser2010} show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski \cite{Ovaskainen2003a} similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial and should thus be conserved.

The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e. growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e. colonization and extinction).
This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape \cite{Hanski2003}, but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition \cite{Becker2015}.
Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
More generally, the dynamics and establishment of environmental reservoirs could be influenced by a number of other factors, such as the length of infectious occupancy or the occupancy history of a patch.
The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.
Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g. by making transmission parameters also function of patch quality.

Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of management traps in wildlife populations currently affected by an environmentally persistent pathogen.
For instance, high quality habitat seems unlikely to form management traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges \cite{Langwig2014, ORegan2015}.
Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units \cite{Conner2004} and the relative importance of local-scale transmission within winter ranges \cite{Farnsworth2006}.
However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist \cite{Stapp2004}.
Indeed, studies have suggested that large colonies are more likely to become infected \cite{Snall2008}, and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.

@@

This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of \cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}.
In particular, for a given pathogen, Hess \cite{Hess1996} similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.  
In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a widespread infection, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in moderate disease spread, increasing movement decreased host occupancy.
These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the three scenarios we outline here based on disease parameters (infectious survival and direct and environmental transmission).
The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g. increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.

Given that the presence of this ecological trap depends on the relative magnitude of pathogen spread, we find that pathogens with long-lived environmental reservoirs only generate traps when infectious survival is low enough that the pathogen is not able to spread widely.
Because environmentally long-lived pathogens are able to spread and persist more easily, increasing longevity can overwhelm the ability of low quality habitat to slow spread and provide refuges for susceptible hosts.
These dynamics reflect the findings of Gog \emph{et al.} \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park \cite{Park2012} expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).
